Exploring dimensions of biodiversity in Japanese ferns

### 2020-07-15 presentations/bsa_2020-07-27/index.Rmd
Update slides
- add example ferns

@@

# Load targets
loadd(list = c(
  "traits_for_dist",
  "ses_div_ferns_spatial",
  "occ_point_data_ferns",
  "protected_areas"),
  cache = ja_fern_cache)

# Change phyloregion cluster to factor
ses_div_ferns_spatial <- ses_div_ferns_spatial %>%
  mutate(taxonomic_cluster = as.factor(taxonomic_cluster) %>% fct_infreq %>% as.numeric %>% as.factor) %>%
  mutate(phylo_cluster = as.factor(phylo_cluster) %>% fct_infreq %>% as.numeric %>% as.factor)

@@

# Save currently used data as back-up in case data produced by plan changes
data <- list(
  ses_div_ferns_spatial = ses_div_ferns_spatial,
  occ_point_data_ferns = occ_point_data_ferns,
  collections_by_country = collections_by_country,
  japan = japan
)

saveRDS(data, here::here("presentations/bsa_2020-07-27/data/data.RDS"))
```

@@

@@

- Diverse flora (ca. 675 spp.)

@@

- Data available for DNA, distribution, and traits

]

@@

```{r tax-bioregion, out.width = "120%", fig.asp = 0.6}
ggplot(ses_div_ferns_spatial, aes(fill = taxonomic_cluster)) +
  geom_sf(color = "transparent") +
  scale_fill_carto_d(name = "Cluster", type = "qualitative", palette = "Bold") +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  map_theme_pp_light() +
  labs(
    title = "Taxonomic bioregions"
  )
```

@@

```{r compare-bioregion, out.width = "120%", fig.asp = 0.7}
a <- ggplot(ses_div_ferns_spatial, aes(fill = taxonomic_cluster)) +
  geom_sf(color = "transparent") +
  scale_fill_carto_d(name = "Cluster", type = "qualitative", palette = "Bold") +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  map_theme_pp_light() +
  labs(
    title = "Taxonomic bioregions"
  ) +
  theme(legend.position = "none")

@@

.pull-right[
Forest types

<img src="images/forest_types_eng.png" alt="drawing" width="280"/>
]

.footnote[`r Citet(bib, "Shimizu2014")`]

>>>>

simple_map_theme_light <- function () {
  theme(
    panel.grid.major = element_line(color = "white", size = 0.1),
    panel.background = element_rect(fill = "grey90"),
    legend.position = "none",
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
}

@@

# Load targets
loadd(list = c(
  "ses_div_ferns_spatial",
  "occ_point_data_ferns",
  "protected_areas",
  "traits_for_dist",
  "comm_ferns",
  "green_list",
  "ppgi"),
  cache = ja_fern_cache)

@@

# Save currently used data as back-up in case data produced by plan changes
data <- list(
  ses_div_ferns_spatial = ses_div_ferns_spatial,
  occ_point_data_ferns = occ_point_data_ferns,
  protected_areas = protected_areas,
  traits_for_dist = traits_for_dist,
  comm_ferns = comm_ferns,
  green_list = green_list,
  ppgi = ppgi,
  collections_by_country = collections_by_country,
  japan = japan
)

saveRDS(data, here::here("presentations/bsa_2020-07-27/data/data.RDS"))

# Change phyloregion cluster to factor
ses_div_ferns_spatial <- ses_div_ferns_spatial %>%
  mutate(taxonomic_cluster = as.factor(taxonomic_cluster) %>% fct_infreq %>% as.numeric %>% as.factor) %>%
  mutate(phylo_cluster = as.factor(phylo_cluster) %>% fct_infreq %>% as.numeric %>% as.factor)
```

@@

```{r count-taxa}
# count number of higher taxa (genera and families)
n_fern_taxa <-
green_list %>%
  mutate(genus = str_split(taxon, "_") %>% map_chr(1)) %>%
  assert(not_na, genus) %>%
  # Add higher-level taxonomy
  left_join(ppgi, by = "genus") %>%
  assert(not_na, class) %>%
  # Filter to only ferns
  filter(class == "Polypodiopsida") %>%
  summarize(
    n_genus = n_distinct(genus),
    n_family = n_distinct(family)
  )
```

@@

- Diverse flora

 - ca. 675 spp.

 - `r n_fern_taxa$n_genus` genera

 - `r n_fern_taxa$n_family` families

@@

- Data available for DNA, distribution, and traits
]

.pull-right[
.center[
<img src="images/ja_example_ferns.png" alt="drawing" height="580"/>
]
]

.footnote[.small[Photos A. Ebihara]]

- Diverse flora

 - ca. 675 spp.

 - `r n_fern_taxa$n_genus` genera

 - `r n_fern_taxa$n_family` families

- Data available for DNA, distribution, and traits
]

@@

```{r tax-bioregion, out.width = "120%", fig.asp = 0.6}
ggplot(ses_div_ferns_spatial, aes(fill = taxonomic_cluster)) +
  geom_sf(color = "transparent") +
  scale_fill_carto_d(name = "Region", type = "qualitative", palette = "Bold") +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  map_theme_pp_light() +
  labs(
    title = "Taxonomic bioregions"
  )
```

```{r find-top-species}
# Extract clusters
clusters <-
  ses_div_ferns_spatial %>%
  as_tibble() %>%
  select(site = "grids", cluster = taxonomic_cluster, -geometry)

# Convert community df to long format
comm_ferns_long <-
  comm_ferns %>%
  rownames_to_column("site") %>%
  pivot_longer(names_to = "taxon", values_to = "abun", -site)

# Find top-scoring species (with high occurrence inside one cluster, and low elsewhere)
top_species <-
  comm_ferns_long %>%
  left_join(clusters, by = "site") %>%
  filter(cluster %in% 1:4) %>%
  group_by(taxon, cluster) %>%
  summarize(
    abun = sum(abun),
    total = n()
  ) %>%
  mutate(
    # percent abundance within a bioregion
    perc_abun = abun / total
  ) %>%
  arrange(taxon, desc(perc_abun)) %>%
  select(taxon, cluster, perc_abun) %>%
  group_by(taxon) %>%
  # take top two highest percent abundances per species, calculate their difference
  slice_max(n = 2, order_by = perc_abun, .preserve = TRUE, with_ties = FALSE) %>%
  mutate(
    top_cluster = cluster[perc_abun == max(perc_abun)]
  ) %>%
  summarize(
    max_dif = max(perc_abun) - min(perc_abun),
    cluster = unique(top_cluster)
  ) %>%
  # keep only those with greater than 65% difference between top 1st and 2nd 
  # most abundant bioregions
  filter(max_dif > 0.5) %>%
  group_by(cluster) %>%
  # Choose top three per cluster for plotting
  slice_max(n = 3, order_by = max_dif, with_ties = FALSE)
```

```{r plot-top-species-prep}
# make helper function to plot top-scoring species
plot_top_taxon <- function (taxon_select, ses_div_ferns_spatial, comm_ferns_long) {
ses_div_ferns_spatial %>%
  left_join(
    filter(comm_ferns_long, taxon == taxon_select),
    by = c(grids = "site")
  ) %>%
  mutate(abun = factor(abun)) %>%
  ggplot(aes(fill = abun)) +
  scale_fill_manual(values = c("#404040", "#d01c8b")) +
  geom_sf(color = "transparent") +
  labs(
    subtitle = str_replace_all(taxon_select, "_", " ")
  ) +
  simple_map_theme_light() +
  theme(
    plot.subtitle = element_text(face = "italic")
  )
}
```

```{r plot-top-species-cluster-1, out.width = "120%", fig.asp = 0.65}
focus_region <-
ses_div_ferns_spatial %>%
  mutate(focus = ifelse(taxonomic_cluster == 1, "yes", "no")) %>%
ggplot(aes(fill = taxonomic_cluster, alpha = focus)) +
  geom_sf(color = "transparent") +
  scale_fill_carto_d(name = "Cluster", type = "qualitative", palette = "Bold") +
  scale_alpha_manual(values = c(yes = 1, no = 0.25)) +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  simple_map_theme_light() +
  labs(subtitle = "Region 1")

focus_region +
plot_top_taxon(top_species$taxon[[1]], ses_div_ferns_spatial, comm_ferns_long) +
plot_top_taxon(top_species$taxon[[2]], ses_div_ferns_spatial, comm_ferns_long) +
plot_top_taxon(top_species$taxon[[3]], ses_div_ferns_spatial, comm_ferns_long) +
plot_layout(ncol = 2, nrow = 2)
```

```{r plot-top-species-cluster-2, out.width = "120%", fig.asp = 0.65}
focus_region <-
ses_div_ferns_spatial %>%
  mutate(focus = ifelse(taxonomic_cluster == 2, "yes", "no")) %>%
ggplot(aes(fill = taxonomic_cluster, alpha = focus)) +
  geom_sf(color = "transparent") +
  scale_fill_carto_d(name = "Cluster", type = "qualitative", palette = "Bold") +
  scale_alpha_manual(values = c(yes = 1, no = 0.25)) +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  simple_map_theme_light() +
  labs(subtitle = "Region 2")

focus_region +
plot_top_taxon(top_species$taxon[[4]], ses_div_ferns_spatial, comm_ferns_long) +
plot_top_taxon(top_species$taxon[[5]], ses_div_ferns_spatial, comm_ferns_long) +
plot_top_taxon(top_species$taxon[[6]], ses_div_ferns_spatial, comm_ferns_long) +
plot_layout(ncol = 2, nrow = 2)
```

```{r plot-top-species-cluster-3, out.width = "120%", fig.asp = 0.65}
focus_region <-
ses_div_ferns_spatial %>%
  mutate(focus = ifelse(taxonomic_cluster == 3, "yes", "no")) %>%
ggplot(aes(fill = taxonomic_cluster, alpha = focus)) +
  geom_sf(color = "transparent") +
  scale_fill_carto_d(name = "Cluster", type = "qualitative", palette = "Bold") +
  scale_alpha_manual(values = c(yes = 1, no = 0.25)) +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  simple_map_theme_light() +
  labs(subtitle = "Region 3")

focus_region +
plot_top_taxon(top_species$taxon[[7]], ses_div_ferns_spatial, comm_ferns_long) +
plot_top_taxon(top_species$taxon[[8]], ses_div_ferns_spatial, comm_ferns_long) +
plot_top_taxon(top_species$taxon[[9]], ses_div_ferns_spatial, comm_ferns_long) +
plot_layout(ncol = 2, nrow = 2)
```

```{r plot-top-species-cluster-4, out.width = "120%", fig.asp = 0.65}
bonin_box <-
ses_div_ferns_spatial %>%
  filter(taxonomic_cluster == 4) %>%
  sf::st_bbox()

expand_value <- 0.4

focus_region <-
ses_div_ferns_spatial %>%
  mutate(focus = ifelse(taxonomic_cluster == 4, "yes", "no")) %>%
ggplot(aes(fill = taxonomic_cluster, alpha = focus)) +
  geom_sf(color = "transparent") +
  scale_fill_carto_d(name = "Cluster", type = "qualitative", palette = "Bold") +
  scale_alpha_manual(values = c(yes = 1, no = 0.25)) +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  simple_map_theme_light() +
  labs(subtitle = "Region 4") +
  annotate(
    "rect", 
           xmin = bonin_box[["xmin"]] - expand_value, 
           xmax = bonin_box[["xmax"]] + expand_value, 
           ymin = bonin_box[["ymin"]] - expand_value, 
           ymax = bonin_box[["ymax"]] + expand_value,
    fill = "transparent",
    color = "grey10")

focus_region +
plot_top_taxon(top_species$taxon[[10]], ses_div_ferns_spatial, comm_ferns_long) +
plot_top_taxon(top_species$taxon[[12]], ses_div_ferns_spatial, comm_ferns_long) +
plot_top_taxon(top_species$taxon[[11]], ses_div_ferns_spatial, comm_ferns_long) +
plot_layout(ncol = 2, nrow = 2)
```

@@

```{r compare-bioregion, out.width = "120%", fig.asp = 0.65}
a <- ggplot(ses_div_ferns_spatial, aes(fill = taxonomic_cluster)) +
  geom_sf(color = "transparent") +
  scale_fill_carto_d(name = "Cluster", type = "qualitative", palette = "Bold") +
  scale_y_continuous(labels = scales::label_number(suffix = "\u00b0N", accuracy = 1)) +
  scale_x_continuous(labels = scales::label_number(suffix = "\u00b0E", accuracy = 1)) +
  map_theme_pp_light() +
  labs(
    title = "Taxonomic bioregions"
  ) +
  theme(legend.position = "none")

@@

.pull-right[
Forest types<br>
.small[`r Citet(bib, "Shimizu2014")`<br><br>]
<img src="images/forest_types_eng.png" alt="drawing" width="280"/>
]
